Confirmation of \srcfull{} as a long period radio transient

### 2025-04-24 paper/main.tex
Udated main.tex to match Overleaf

@@

@@

@@

\section{Introduction} \label{sec:intro}

\srcfull{} (hereafter, \src{}) is a transient radio source discovered in the Variables and Slow Transients (VAST) survey \citep{2024MNRAS.535..909D}.
A single, highly polarized pulse was observed, lasting approximately 2 minutes, exhibiting a steep spectrum ($\alpha = -3.1$).
Extensive follow up in new and archival data at multiple wavelengths turned up no other pulses.
Despite this, the authors concluded that \src{} was most likely a long period transient (LPT), a newly emerging class of radio source with periods ranging from minutes to hours (see Table \ref{tbl:lpts} for an up-to-date list).

\begin{deluxetable*}{lcl}
\tablecaption{Known LPTs, arranged in ascending order of period\label{tbl:lpts}}
\tablehead{
\colhead{LPT} & \colhead{Period} & \colhead{References} \\ 
 & \colhead{(min)} & 
} 
\startdata
%AR Scorpii (J162147.28-225310.3) & 1.97 &  \\
%ASKAP J175154.89-255135.3 & ${\sim}$3.6 &  \\
%J1912-4410 & 5.3 &  \\
CHIME J0630+25             & 7.02 & \citet{2024arXiv240707480D} \\
GLEAM-X J162759.5-523504.3 & 18.2 & \citet{2022Natur.601..526H} \\
GPM J1839-10               & 22.0 & \citet{2023Natur.619..487H} \\
ASKAP J1832-0911           & 44.3 & \todo{Wang} \\
ASKAP J193505.1+214841.0   & 53.8 & \citet{2024NatAs...8.1159C} \\
ASKAP J175534.9-252749.1   & 69.8 & \citet{2024MNRAS.535..909D} \\
GCRT J1745-3009            & 77.0 & \citet{2005Natur.434...50H} \\
ILT J1101+5521             & 126  & \citet{2024arXiv240811536D} \\
GLEAM-X J0704-37           & 175  & \citet{2024arXiv240815757H} \\
ASKAP J1839-0756           & 387  & \todo{Lee} \\
\enddata
%\tablecomments{This is a sample table. Add more rows as needed, and include notes in this area.}
\end{deluxetable*}

@@

[Acknowledgements go here\dots]

>>>>

\usepackage{amsmath}

@@

\defcitealias{2024MNRAS.535..909D}{Paper~I}

@@

\author[0009-0003-0996-9176]{Csanad Horv\'{a}th}
\affiliation{International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia}

\author{Angie Waszewski}
\affiliation{International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia}

\author{John Morgan}
\affiliation{International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia}

@@

\section{Introduction} \label{sec:introduction}

\srcfull{} (hereafter, \src{}) is a transient radio source discovered in the Variables and Slow Transients (VAST) survey \citep[][hereafter \citetalias{2024MNRAS.535..909D}]{2024MNRAS.535..909D}.
A single, highly polarized pulse was observed, lasting approximately 2 minutes, exhibiting a steep spectrum ($\alpha = -3.1$).
Extensive follow up in new and archival data at multiple wavelengths turned up no other pulses.
Despite this, the authors concluded that \src{} was most likely a long period transient (LPT), a newly emerging class of radio source with periods ranging from minutes to hours (see Table \ref{tbl:lpts} for an up-to-date list).

Although optical follow-up has confirmed that at least some LPTs are white dwarfs with main sequence companions \citep{2024arXiv240811536D,2025arXiv250103315R}, it remains unclear if this model can account for all known LPTs \citep[e.g.][]{2022ApJ...940...72R,Lee2025}.
Unfortunately, optical follow-up is not possible for LPTs and LPT candidates which are obscured by Galactic foregrounds.
This is almost certainly the case for \src{}, which is situated very close to the Galactic plane ($b = -0\overset{\circ}{.}12$), and has a dispersion measure (DM) derived distance of ${\sim}4.7\,$kpc \citepalias{2024MNRAS.535..909D}; unsurprisingly, optical follow-up at \src{}'s location did not turn up any source down to ${\sim}23$ AB mag.
Without more detections in the radio or at other wavelengths, it seemed unlikely that the nature of \src{} could ever be resolved.

In this short paper, we present new radio observations (spanning frequencies from $154\,$MHz to $1.7\,$GHz) in which \src{} was re-detected, confirming that it is indeed an LPT with a period of ${\sim}1.16\,$hours.
In \S\ref{sec:observations}, we describe the observations themselves, some of which were taken as part of independent surveys, and some of which were taken as dedicated follow-up of \src{} once it became apparent that it was still active.
In \S\ref{sec:analysis} we present a timing analysis of the new detections, as well as revised estimates of the DM and the scattering timescale.
We also include new polarization detections in which the polarisation angle (PA) behavior differs significantly from the originally detected pulse.
In \S\ref{sec:discussion} we revisit the question of the nature of \src{}, arguing in favor of a white dwarf interpretation.
The main conclusions are summarized in \S\ref{sec:conclusions}.

\begin{deluxetable*}{lcl}
\tablecaption{Known and candidate LPTs, arranged in ascending order of period\label{tbl:lpts}}
\tablehead{
\colhead{LPT} & \colhead{Period} & \colhead{References} \\
 & \colhead{(min)} &
}
\startdata
%AR Scorpii (J162147.28-225310.3) & 1.97 &  \\
%ASKAP J175154.89-255135.3 & ${\sim}$3.6 &  \\
%J1912-4410 & 5.3 &  \\
CHIME J0630+25             & 7.02 & \citet{2024arXiv240707480D} \\
GLEAM-X J162759.5-523504.3 & 18.2 & \citet{2022Natur.601..526H} \\
GPM J1839-10               & 22.0 & \citet{2023Natur.619..487H} \\
ASKAP J1832-0911           & 44.3 & \todo{Wang} \\
ASKAP J193505.1+214841.0   & 53.8 & \citet{2024NatAs...8.1159C} \\
ASKAP J175534.9-252749.1   & 69.8 & \citet{2024MNRAS.535..909D} \\
GCRT J1745-3009            & 77.0 & \citet{2005Natur.434...50H} \\
ILT J1101+5521             & 126  & \citet{2024arXiv240811536D} \\
GLEAM-X J0704-37           & 175  & \citet{2024arXiv240815757H} \\
ASKAP J1839-0756           & 387  & \citet{Lee2025} \\
\enddata
\end{deluxetable*}

\section{Observations} \label{sec:observations}

The observations included in this paper were made at the Murchison Widefield Array \citep[MWA;][]{Tingay2013},

\subsection{MWA} \label{sec:mwa}

\subsection{ASKAP} \label{sec:askap}

\subsection{MeerKAT} \label{sec:meerkat}

%\begin{deluxetable*}{cccccc}
%\tablecaption{Pulse detections and non-detections}
%\tablehead{
%\colhead{Obs start} & \colhead{Duration} & \colhead{Telescope} & \colhead{$\nu$} & %\colhead{$\Delta\nu$} & \colhead{Detected?} \\
%\colhead{time (UTC)} & \colhead{(s)} & & \colhead{(MHz)} & \colhead{(MHz)} &
%}
%\startdata
%2024-09-19 09:25:02 & 592\tablenotemark{a} & MWA & 184.96 & 30.72 & Y \\ % 1410773120 and 1410773416
%2024-09-19 10:34:06 & 592\tablenotemark{a} & MWA & 184.96 & 30.72 & Y \\ % 1410777264 and 1410777560
%2024-09-19 12:57:10 & 296 & MWA & 184.96 & 30.72 & Y \\ % 1410785848
%2024-09-19 14:06:14 & 296 & MWA & 184.96 & 30.72 & Y \\ % 1410789992
%2024-10-09 10:55:50 & 296 & MWA & 200.32 & 30.72 & Y \\ % 1412506568
%2024-10-09 12:05:34 & 296 & MWA & 200.32 & 30.72 & Y \\ % 1412510752
%2024-10-10 11:21:10 & 296 & MWA & 200.32 & 30.72 & Y \\ % 1412594488
%2024-10-10 12:30:54 & 296 & MWA\tablenotemark{b} & 200.32 & 30.72 & Y \\ % 1412598672
%2024-10-11 10:36:46 & 296 & MWA & 200.32 & 30.72 & Y \\ % 1412678224
%2024-10-11 11:46:30 & 296 & MWA & 200.32 & 30.72 & Y \\ % 1412682408
%2024-10-12 11:02:06 & 296 & MWA & 200.32 & 30.72 & Y \\ % 1412766144
%2024-10-12 12:11:50 & 296 & MWA & 200.32 & 30.72 & Y \\ % 1412770328
%2024-10-13 10:17:42 & 296 & MWA & 200.32 & 30.72 & Y \\ % 1412849880
%2024-10-13 11:27:26 & 296 & MWA & 200.32 & 30.72 & Y \\ % 1412854064
%2024-10-14 10:43:02 & 296 & MWA & 200.32 & 30.72 & Y \\ % 1412937800
%2024-10-14 11:52:46 & 296 & MWA & 200.32 & 30.72 & Y \\ % 1412941984
%2024-10-16 10:23:58 & 296 & MWA & 200.32 & 30.72 & Y \\ % 1413109456
%2024-10-16 11:33:42 & 296 & MWA & 200.32 & 30.72 & Y \\ % 1413113640
%2024-10-17 11:59:02 & 296 & MWA & 200.32 & 30.72 & Y \\ % 1413201560
%2024-10-19 10:30:14 & 296 & MWA & 200.32 & 30.72 & Y \\ % 1413369032
%2024-10-19 11:39:58 & 296 & MWA & 200.32 & 30.72 & Y \\ % 1413373216
%2024-10-19 13:54:36 & 601 & MeerKAT\tablenotemark{c} & 812.8 & 247.6 & Y \\ % 1413381294
%\enddata
%\tablenotetext{a}{Two contiguous 296-second MWA observations were combined}
%\tablenotetext{b}{Too much RFI to use in downstream analysis}
%\tablenotetext{c}{Two components in pulse}
%\end{deluxetable*}

\begin{deluxetable*}{cccccc}
\tablecaption{Observations}
\tablehead{
\colhead{Date range} & \colhead{Telescope} & Project & \colhead{$\nu$} & \colhead{$\Delta\nu$} & \colhead{Number of} \\
\colhead{(MJD)} & & & \colhead{(MHz)} & \colhead{(MHz)} & \colhead{pulses}
}
\startdata
59965 & ASKAP\tablenotemark{a} & SB47253 & 887.5 & 288 & 1 \\
59966 & MWA & G0060 & 161.92 & 15.36 & 1 \\
60084 & MWA & G0057 & 154.24 & 30.72 & $\le$ 1 \\
60092 - 60093 & MeerKAT & & & & 4 \\
60463 - 60563 & MWA & G0080 & 200.32 & 30.72 & 18 \\
60503 & ASKAP &  & 887.5 & 288 & 1 \\
60572 & MWA & D0042 & 184.96 & 30.72 & 4 \\
60592 - 60602 & MWA & D0042 & 200.32 & 30.72 & 16 \\
60602 & MeerKAT & & 812.8 & 495.3 & 1 \\
\enddata
\tablenotetext{a}{Original discovery \citep{2024MNRAS.535..909D}}
\end{deluxetable*}

\section{Analysis \& results} \label{sec:analysis}

[Reference for scattering kernels:] \citep{10.1093/mnras/stw1724}.

\subsection{Pulse morphology}

All observed pulses consist of a single burst, approximately Gaussian in shape, with the sole exception of the MeerKAT pulse observed on 2024-10-19, for which two bursts were observed separated by ${\sim}125\,$s.
Generally, the MWA pulses tend
\begin{itemize}
    \item to be wider than their higher frequency counterparts,
    \item to exhibit an asymmetry similar to scattering, and
    \item to be very similar in morphology to each other.
\end{itemize}

The asymmetry is consistent with a scattering time scale of a few tens of seconds at MWA frequencies, which scales to a few tens of milliseconds at 1\,GHz.
This is in broad agreement with the model predictions of NE2001 ($\tau_{\rm sc,1 GHz} = 0.066\,$s) and YMW16 ($\tau_{\rm sc,1 GHz} = 0.075\,$s) at the source's location assuming a DM of 710\,pc\,cm$^{-3}$.
The effect of scattering on the timing is discussed in \S\ref{sec:timing} below.

\subsection{Timing solution} \label{sec:timing}

%Following the positive identification of several pulses in the GPM data set, we performed a grid search of periods that would result in an integer number of pulses between two widely spaced pulses.
%A candidate period of $P = 4186.4 \pm 0.1$ was found for which all detections fell within a narrow window of ${\sim}0.05$ of phase.
% We further refined this by attempting to phase connect to Dougal's original pulse, and got $4186.33 \pm 0.02$ s, but this needn't be part of the story, since even the above was enough to motivate the MWA campaign that followed.
%This motivated the follow-up MWA observations (under project code D0042) in which two consecutive pulses ${\sim}4190$ seconds apart were detected without any discernible emission between them, ruling out periods $P/n$, for $n \ge 2$.
%The lever arm afforded by both the GPM and D0042 detections proved sufficient to phase-connect to the original pulse published by \citet{2024MNRAS.535..909D}, yielding a refined period of $P = 4186.33 \pm 0.02$~s.

%We then measured times of arrival (ToAs) of individual pulses.
%The ASKAP and MWA pulses were morphologically similar and well fit by a simple Gaussian.

\subsection{Scattering}

\begin{figure}[th]
    \centering
    \includegraphics[width=0.95\linewidth]{stacked_spectra.pdf}
    \caption{Stacked (semi-transparent) dynamic spectra, where each rectangle indicates the observing campaign whose spectra were barycentred and folded according to the ephemeris. The spectra have not been dedispersed, but the cyan dashed lines indicate where the two components seen in the MeerKAT (2024) observation would appear due to dispersion. The white curve in the top panel is the dedispersed profile of the MeerKAT (2024) observation, and the white curves in the bottom panel are the same profile subjected to scattering with $\tau_{\rm sc} = \tau_{\rm sc,1\,GHz} (\nu/{\rm 1\,GHz})^{-4}$ at frequencies 175, 195, and 210 MHz, where $\tau_{\rm sc,1\,GHz} = 0.07\,$s.}
    \label{fig:stacked_spectra}
\end{figure}

\begin{figure*}[pt]
    \centering
    \includegraphics[width=0.9\linewidth]{pulsestack.pdf}
    \caption{Pulsestack of barycentred, dedispersed profiles from the ephemeris. The dashed red vertical line marks zero phase. Baselines have been fitted and subtracted from each lightcurve.}
    \label{fig:pulsestack}
\end{figure*}

\subsection{Polarization} \label{sec:polarization}

\section{Discussion} \label{sec:discussion}

\begin{enumerate}
    \item Why does the source appear to not be active all the time at higher frequencies?
    \item As per convo with John (see MWA Users Slack, 2025-01-24), point out that (lack of detection of source in) MWA IPS is not very constraining in terms of scattering disk / timescale / screen.
\end{enumerate}

\section{Conclusions} \label{sec:conclusions}

@@

% PTUSE (https://skaafrica.atlassian.net/wiki/spaces/ESDKB/pages/1591672833/User+Supplied+Equipment+USE#PTUSE)
Observations made use of the Pulsar Timing User Supplied Equipment (PTUSE) servers at MeerKAT which were funded by the MeerTime Collaboration members ASTRON, AUT, CSIRO, ICRAR-Curtin, MPIfR, INAF, NRAO, Swinburne University of Technology, the University of Oxford, UBC and the University of Manchester.  The system design and integration was led by Swinburne University of Technology and Auckland University of Technology in collaboration with SARAO and supported by the ARC Centre of Excellence for Gravitational Wave Discovery (OzGrav) under grant CE170100004.
